Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-03 talk/talk.tex
Fiddle with talk, and add some stuff.

\documentclass{beamer}

@@

@@

\vspace{-1mm}
\begin{itemize}
\item a glacier which is in balance with its climate
    \begin{itemize}
    \item the simplest such problem (grounded, nonsliding, isothermal)
    \item re~\emph{steady}: solve this and you can handle implicit steps too
    \end{itemize}
\item steady climate data given on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item determine: surface elevation $s(x,y)$ on $\Omega$, domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{block}{steady surface kinematical equation (SKE)}
$$\bu|_s \cdot \bn_s + a=0$$
\end{block}

\begin{itemize}
\item $\bn_s = \left<-s_x,-s_y,1\right>$  \quad (normal to ice surface; upward)
\item \emph{on} the ice surface, in steady-state there is balance (SKE) between motion and climatic mass balance
\item \emph{off} the ice we know $a\le 0$ in steady state
\item for all of this talk:
    \begin{itemize}
    \item $s$ is assumed to be well-defined (no overhangs)
    \item $s=b$ where ice is not present, so $s$ is defined on all of $\Omega$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the SIGP is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}

@@

    \begin{itemize}
    \item plus Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \end{itemize}
\item SIGP is a free-boundary problem for $s$ defined on $\Omega$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the SIGP directly, or doing implicit time-stepping
    \begin{itemize}
    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} \dots nonscalable, w.~``fake ice''
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation} (below)
    \end{itemize}
\end{itemize}

\medskip
\begin{columns}
    \column{0.8\textwidth}
        \hfill \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.2\textwidth}
        \scriptsize \emph{show clip of PISM run by Julien Seguinot}
\end{columns}
\end{frame}

@@

\begin{block}{ice dynamics operator}
$\Phi : s \mapsto \bus \cdot \bn_s$

@@

\begin{itemize}
\item to compute $\Phi(s)$, solve Stokes (weak form) problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0$$
over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ and extract trace $\bus$

\medskip
    \begin{itemize}
    \item optimal solver already exists for this problem (Isaac et al \cite{IsaacStadlerGhattas2015})
    \end{itemize}
\item extend by zero so $\Phi(s)$ is defined on all of $\Omega$
\end{itemize}

@@

\item well-known fact: \quad  NCP (strong form) $\leftrightarrow$ VI (weak form)
\item VI weak form: find $s \in \mathcal{K} = \{r \ge b\}$ so that
    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
for all $r \in \mathcal{K}$, where $F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$

\medskip
    \begin{itemize}
    \item whether this VI is well-posed is open, but the SIA version has an existence proof (Jouvet \& Bueler \cite{JouvetBueler2012})
    \end{itemize}
\end{itemize}

@@

\begin{block}{numerical goal}
Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.  Solve VI

$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$

to give admissible $s^h \in \mathcal{K}^h$.  Use the solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$ (and thus $F(s^h)$).
\end{block}

\begin{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
    \end{itemize}
\item optimal requires multigrid
\item<2> \emph{for now these are aspirations only, but let's give it a shot}
\end{itemize}
\end{frame}

\begin{frame}{new method}

@@

\includegraphics[width=0.8\textwidth]{figs/extruded.png}
\end{frame}

\begin{frame}{extra}
 NCP is a barrier to exact numerical mass conservation \cite{Bueler2021conservation}
\end{frame}

@@

\begin{frame}{References}

%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

\begin{thebibliography}{1}
\bibitem{Bueler2021conservation}
E.~Bueler. \textbf{Conservation laws for free-boundary fluid layers.} {\em SIAM J. Appl. Math.}, to appear.

\bibitem{BuelerMitchell2022}
E.~Bueler and L.~Mitchell. \textbf{Multilevel computation of glacier geometry from Stokes dynamics.} in preparation

\bibitem{IsaacStadlerGhattas2015}
T.~Isaac, G.~Stadler, and O.~Ghattas.  \textbf{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics.} {\em SIAM J. Sci. Comput.}, 37(6):B804--B833, 2015.

@@

\bibitem{WirbelJarosch2020}
A.~Wirbel and A.~H. Jarosch. \textbf{Inequality-constrained free-surface evolution in a full {S}tokes ice flow model (\textit{evolve\_glacier v1.1}).} {\em Geoscientific Model Development}, 13(12):6425--6445, 2020.
\end{thebibliography}

\end{document}

>>>>

\documentclass[usepdftitle=false]{beamer}

@@

\usepackage{hyperref}
\hypersetup{
    pdftitle={Evolution of ice sheet geometry using Stokes dynamics}
}

@@

\vspace{-1mm}
\begin{itemize}
\item a glacier which is in balance with its climate
    \begin{itemize}
    \item the simplest such problem (grounded, nonsliding, isothermal)
    \item re~\emph{steady}: solve this and you can handle implicit steps too
    \end{itemize}
\item steady data given on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item determine: surface elevation $s(x,y)$ on $\Omega$, domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow, so there is an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
    $$\bu|_s \cdot \bn_s + a=0 \qquad \text{\emph{steady state}}$$

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \end{itemize}
\item \emph{off} the ice we know $a\le 0$ in steady state
\end{itemize}

\bigskip
\begin{columns}
    \column{0.6\textwidth}
        \begin{itemize}
        \item for all of this talk:
            \begin{itemize}
            \item $s$ is assumed to be well-defined (no overhangs)
            \item $s=b$ where ice is not present, so $s$ is defined on all of $\Omega$
            \end{itemize}
        \end{itemize}
    \column{0.4\textwidth}
        \includegraphics[width=1.05\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item the SIGP free-boundary problem determines $s$ on $\Omega$
\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}

@@

    \begin{itemize}
    \item plus Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \item \emph{count the 3 nonlinearities}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the SIGP directly
    \begin{itemize}
    \item or doing implicit time-stepping (\emph{basically the same as SIGP})
    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} (\emph{semi-coupled, nonscalable, artificial ice layer})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item mostly with shallow approximations, e.g.~PISM run below
    \end{itemize}
\end{itemize}

\medskip
\begin{columns}
    \column{0.8\textwidth}
        \hfill \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.2\textwidth}
        \scriptsize \emph{show clip of Alpine run by Julien Seguinot}
\end{columns}
\end{frame}

@@

\begin{block}{definition: ice dynamics operator}
$\Phi : s \mapsto \bus \cdot \bn_s$

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0$$

    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz \cite{JouvetRappaz2011})
    \item optimal solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
    \item extract trace $\bus$, then $\Phi(s) = \bus \cdot \bn_s$, extend by zero to all of $\Omega$
    \end{itemize}
\end{itemize}

@@

\item NCP (strong form) $\leftrightarrow$ VI (weak form)

\begin{block}{SIGP weak form}
find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
for all $r \in \mathcal{K}$, where $F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$
\end{block}

\item well-posedness is open
    \begin{itemize}
    \item existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
    \end{itemize}

\end{itemize}

@@

\begin{block}{numerical goal}
Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.  Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$.  Solve VI for $s^h \in \mathcal{K}^h$:

$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
\end{block}

\begin{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
        \begin{itemize}
        \item[{\color{black} $\circ$}] optimal requires multigrid
        \end{itemize}
    \item<2> \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{what thing is $\Phi$?}

\begin{block}{SIGP weak form}
find $s \in \mathcal{K}$ so $F(s)[r - s] \ge \ip{a}{r-s}$ for all $r \in \mathcal{K}$
\end{block}

\begin{itemize}
\item $F(s)[w] = - \int_\Omega \Phi(s)\, w$ is basically dual of $\Phi(s)=\bus\cdot\bn_s$
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item some kind of integral operator (\emph{convolve with Stokeslets})
    \item $\Phi$ is concentrated near the diagonal, i.e.~short range
    \end{itemize}
\item yet it is a differential operator in the SIA case:
\small
    $$\Phi_{\text{SIA}}(s) = \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} + \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
\normalsize
\end{itemize}
\end{frame}

\begin{frame}{multilevel constraint decomposition}

@@

\includegraphics[width=0.8\textwidth]{figs/extruded.png}

\input{tikz/vcycle.tex}
\end{frame}

@@

\begin{frame}{References}
%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

\begin{thebibliography}{1}
\bibitem{BuelerMitchell2022}
E.~Bueler and L.~Mitchell. \textbf{Multilevel computation of glacier geometry from Stokes dynamics.} in preparation

\bibitem{IsaacStadlerGhattas2015}
T.~Isaac, G.~Stadler, and O.~Ghattas.  \textbf{Solution of nonlinear Stokes equations \dots with application to ice sheet dynamics.} {\em SIAM J. Sci. Comput.}, 37(6):B804--B833, 2015.

@@

\bibitem{JouvetRappaz2011}
G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary Stokes problem arising in glaciology.} {\em Adv. Numer. Anal.}, 2011.

\bibitem{WirbelJarosch2020}
A.~Wirbel and A.~H. Jarosch. \textbf{Inequality-constrained free-surface evolution in a full {S}tokes ice flow model \dots} {\em Geosci. Model Dev.}, 13(12):6425--6445, 2020.
\end{thebibliography}
\end{frame}

\begin{frame}{extra}

NCP is a barrier to exact numerical mass conservation\footnote{E.~Bueler. \textbf{Conservation laws for free-boundary fluid layers.} \em{SIAM J.~Appl.~Math.}, to appear.}
\end{frame}
\end{document}
